---
title: "Glioma Brain Tumour Analysis"
author: "Chigozie Nkwocha"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(limma))
suppressMessages(library(tidyheatmaps))
```


```{r}
winsorize_within_class <- function(df, groups=NULL, percent=90){
  df_copy <- df
  
  winsorized_df <- data.frame()
  
  cutoff <- round((1-percent/100)/2,3)
  if (!is.null(groups)){
    uniq_groups <- unique(groups)
    for (cls in uniq_groups){
      # get group indices
      group_idx <- which(groups == cls)
      
      # subset group
      group_df <- df_copy[, group_idx]
      
      # get quantiles of each gene
      upper <- apply(group_df, 1, quantile, probs=1-cutoff)
      lower <- apply(group_df, 1, quantile, probs=cutoff)
      
      # Winsorize
      group_df <- apply(group_df, 2, function(x) ifelse(
        x < lower, lower, ifelse(x > upper, upper, x)))
      
      group_df <- as.data.frame(group_df) 
      # transpose to make genes as columns and samples as rows for easy appending to winsorised_df
      winsorized_df <- rbind(winsorized_df, t(group_df) |> as.data.frame())
    }
    
    # convert back with genes as rows and samples as columns
    winsorized_df <- winsorized_df |> t() |> as.data.frame()
    return(winsorized_df)
  }
  else{
    # get quantiles of each gene
    upper <- apply(df_copy, 1, quantile, probs=1-cutoff)
    lower <- apply(df_copy, 1, quantile, probs=cutoff)
    
    # Winsorize
    df_copy <- apply(df_copy, 2, function(x) ifelse(
      x < lower, lower, ifelse(x > upper, upper, x)))
    df_copy <- as.data.frame(df_copy)
  }
  return(df_copy)
}
```


```{r}
read_expr_data <- function(file, gene_id_file){
  expr_df <- read.csv(file, row.names = 1)
  gene_df <- read.csv(paste0('../microarray_genes_ids/', gene_id_file))
  
  # transform data to log2
  max_val <- max(expr_df)
  
  # if max_val above 20 (log), then it's probably not log2 transformed
  if (max_val > 20){
    expr_df <- log2(1+expr_df)
  }
  
  # map probe IDs to gene symbols
  expr_df <- expr_df |>
    rownames_to_column('ID_REF') |>
    merge(gene_df, by.x='ID_REF', by.y='ID') |>
    select(Genes, everything(), -ID_REF)
  
  # get the max values of genes with more than one probes
  expr_df <- expr_df |>
    group_by(Genes) |> 
    summarise(across(everything(), max)) |> 
    column_to_rownames('Genes')
  
  # quantile normalize expression data
  expr_df <- normalizeBetweenArrays(expr_df, method='quantile')
  
  # return with genes as rows and samples as columns
  return(expr_df)
}
```


```{r}
# get gene expression and gene ids files and remove miRNA files
expr_files <- list.files(pattern='*_exprs*.csv')
metadata_files <- list.files(pattern='*_metadata*.csv')


gene_id_files <- list.files('../microarray_genes_ids/', pattern = '^GSE.+.csv')

# remove mirna data
expr_files <- expr_files[str_detect(expr_files, '^(?!.+mirna).*')]
metadata_files <- metadata_files[str_detect(metadata_files, '^(?!.+mirna).*')]
gene_id_files <- gene_id_files[str_detect(gene_id_files, '^(?!GSE.+_GPL21572).*')]

c(length(expr_files), length(gene_id_files), length(metadata_files))
```


```{r}
# merging expression and metadata from each batch experiment
winsorize <- TRUE
expr_data <- data.frame()
meta_data <- data.frame()
batches <- c()

pb <- progress::progress_bar$new(
  format = "[:bar] :percent :elapsed",
  total = length(expr_files),
  width = 60
)

for (i in 1:length(expr_files)){
  expr_file <- expr_files[i]
  gse_id <- str_extract(expr_file, 'GSE\\d+')
  gene_file <- gene_id_files[grepl(gse_id, gene_id_files)]
  metadata_file <- metadata_files[grepl(gse_id, metadata_files)]

  
  expr_df <- read_expr_data(expr_file, gene_file)
  
  meta_df <- read_csv(metadata_file, show_col_types = FALSE,
                      col_select = c('sample_id', 'tumor_class', 'tumor_grade'))
  meta_df['gse_id'] <- gse_id
  
  # treat outliers
  # if winsorize, winsorize else move samples to rows and genes to columns 
  # to append samples from other datasets
  if (winsorize){
    expr_df <- winsorize_within_class(expr_df, replace_na(meta_df$tumor_grade, 'Normal'))
    expr_df <- t(expr_df) |> as.data.frame()
  } else{
    expr_df <- t(expr_df) |> as.data.frame()
  }
  
  
  batches <- c(batches, rep(gse_id, nrow(expr_df)))
  expr_data <- rbind(expr_data, expr_df)
  meta_data <- rbind(meta_data, meta_df)
  
  
  pb$tick()
}

rm(meta_df, expr_df)
expr_data <- t(expr_data) |> as.data.frame()
```

## Selecting only Glioma related tumours

Here, we will be selecting glioma related tumours: ependymoma, astrocytoma, glioblastoma, oligoastrocytoma, oligodendroglioma
```{r}
gliomas <-  'ependymoma|astrocytoma|glioblastoma|gbm|oligoastrocytoma|oligodendroglioma|oligodendrocytoma|high.+grade.+glioma|normal|non-tumor'
# get rows of metadata that are glioma cells and update batches with it
is_glioma <- str_detect(
  str_to_lower(replace_na(meta_data$tumor_class, 'missing')), 
  gliomas, negate=F)

batches <- batches[is_glioma]
```


```{r}
meta_data <- meta_data |>
  filter(str_detect(
    str_to_lower(tumor_class), 
    gliomas, negate=F))
```

```{r}
# select samples in metadata
expr_data <- expr_data[, meta_data$sample_id]
```


```{r}
meta_data <- meta_data |>
  mutate(tumor_grade = str_to_lower(tumor_grade),
         tumor_class = str_to_lower(tumor_class)) |>
  mutate(tumor_class = str_replace(tumor_class, '.*oligoastrocytoma', 'mixed gliomas'),
         tumor_class = str_replace(tumor_class, '.*[aA]strocytoma', 'astrocytoma'),
         tumor_class = str_replace(tumor_class, 'gbm|glioblas.+', 'glioblastoma'),
         tumor_class = str_replace(tumor_class, 'non-tumor', 'normal'), 
         tumor_class = str_replace(tumor_class, '.*[oO]ligodendrocytoma|anaplastic oligodendroglioma', 
                                   'oligodendroglioma'))
```


```{r}
meta_data <- meta_data |>
  # convert to lower and rename
  mutate(tumor_grade = case_match(tumor_grade,
                                  'grade i' ~ 'G1', 'grade ii' ~ 'G2',
                                  'grade iii' ~ 'G3', 'grade iv' ~ 'G4',
                                  .default = tumor_grade),
         tumor_grade = str_replace(tumor_grade, 'grade\\s', 'G'),
         tumor_grade = replace_na(tumor_grade, 'Normal'),
         tumor_class = str_replace(tumor_class, 'non-tumor', 'normal')) |>
  # tumour type
  mutate(tumor_type = case_when(
    tumor_grade %in% c('G3', 'G4') ~ 'malignant', 
    tumor_grade %in% c('G1', 'G2') ~ 'benign', 
    TRUE ~ tumor_grade))
```


```{r}
table(meta_data$tumor_grade)
table(meta_data$tumor_type)
table(meta_data$tumor_class)
```


```{r}
# get gene names
genes <- rownames(expr_data)
```

```{r}
# let's check the minumum value of an expression level
min(expr_data); max(expr_data)
```

From above, there are expression levels whose values are below zero. It could be that one was not added to the original values before log transformation. Hence, to make log values between 0 and the max value, we will retransform back to the original values and log transform again.

```{r}
expr_data <- log2(1+2^expr_data)
```


```{r}
min(expr_data); max(expr_data)
```




## Checking for batch effect

Next we will perform batch effect correction to remove any technical variability that can affect our analysis. Variability could be due to difference in experimental conditions etc. Since our datasets comes from different experiments, we will perform batch effect correction. Firstly, we will need to confirm if we need to do batch effect correction. We will perform a PCA on the gene expression data but we will transpose our data to let the samples become rows and genes columns. After PCA, we will visualise the first two principal components and color the points by the batch IDs (GEO ID).
```{r}
pca <- prcomp(t(expr_data), scale. = T)

exp.var  <- round(100*pca$sdev^2/sum(pca$sdev^2),1)

ggplot(pca$x, aes(PC1, PC2, color=factor(meta_data$gse_id))) +
  geom_point() +
  theme_light() +
  theme(plot.title = element_text(face='bold')) +
  labs(title='PCA for microarray data (Before Batch correction)', 
       x = paste0('PC1 (', exp.var[1], '%)'), 
       y=paste0('PC2 (', exp.var[2], '%)'),
       color='GSE ID')
```


The figure shows that there's some variability in some of the datasets. This variability will obscure our analysis. Hence, we need to perform batch effect normalisation. We see that the first principal component account for most of the variability (over 50%) in the datasets.


```{r}
# remove batch effects
expr_data <- removeBatchEffect(expr_data, batches, group=meta_data$tumor_grade) |>
  as.data.frame()
```

```{r}
pca <- prcomp(t(expr_data), scale. = T)

exp.var  <- round(100*pca$sdev^2/sum(pca$sdev^2),1)

ggplot(pca$x, aes(PC1, PC2, color=factor(meta_data$gse_id))) +
  geom_point() +
  theme_light() +
  theme(plot.title = element_text(face='bold')) +
  labs(title='PCA for microarray data (After Batch correction)', 
       x = paste0('PC1 (', exp.var[1], '%)'), 
       y=paste0('PC2 (', exp.var[2], '%)'),
       color='GSE ID')
```

We see that we have successfully removed the batch effects

```{r}
write.csv(expr_data, 'glioma_cancer_exprs.csv', row.names = T)
write.csv(meta_data, 'glioma_cancer_metadata.csv', row.names = F)
```


## Analysis

Here, we will perform some exploratory analysis on our gene expression data. Firstly, we will determine the gene with the top 10 highest and lowest expression levels across all the samples.
```{r}
# Highest and least expressing genes
mean_expressions <- rowMeans(expr_data)

top10 <- sort(mean_expressions, decreasing = T)[1:10]
least10 <- sort(mean_expressions, decreasing = F)[1:10]
```

```{r}
top10
least10
```


```{r fig.width=9}
expr_data[names(top10),] |>
  rownames_to_column('Genes') |>
  pivot_longer(-Genes, names_to = 'samples', values_to = 'expression_levels') |>
  merge(meta_data[c('sample_id', 'tumor_grade')], by.x='samples', by.y='sample_id') |>
  mutate(tumor_grade = replace_na(tumor_grade, 'Normal')) |>
  ggplot(aes(x=tumor_grade, y=expression_levels, fill=tumor_grade)) +
  geom_boxplot(alpha=0.6) +
  facet_wrap(~Genes, nrow=2, scales='free_y') +
  theme_light() +
  theme(legend.position = 'none', 
        panel.grid = element_blank(),
        axis.title = element_text(size=11),
        axis.text = element_text(size=8),
        strip.background = element_rect(fill='steelblue'),
        strip.text = element_text(color='white', face='bold'),
        plot.title = element_text(face='bold', size=13)) +
  labs(title='Top 10 Genes by expression levels', 
       y='Expression Levels', x='Tumor Grade')
```

```{r fig.width=9}
expr_data[names(least10),] |>
  rownames_to_column('Genes') |>
  pivot_longer(-Genes, names_to = 'samples', values_to = 'expression_levels') |>
  merge(meta_data[c('sample_id', 'tumor_grade')], by.x='samples', by.y='sample_id') |>
  mutate(tumor_grade = replace_na(tumor_grade, 'Normal')) |>
  ggplot(aes(x=tumor_grade, y=expression_levels, fill=tumor_grade)) +
  geom_boxplot(alpha=0.6) +
  facet_wrap(~Genes, nrow=2, scales='free_y') +
  theme_light() +
  theme(legend.position = 'none', 
        panel.grid = element_blank(),
        axis.title = element_text(size=11),
        axis.text = element_text(size=8),
        strip.background = element_rect(fill='steelblue'),
        strip.text = element_text(color='white', face='bold'),
        plot.title = element_text(face='bold', size=13)) +
  labs(title='Least 10 Genes by expression levels', 
       y='Expression Levels', x='Tumor Grade')
```


```{r}
# PCA by tumour group
ggplot(pca$x, aes(PC1, PC2, color=factor(meta_data$tumor_grade))) +
  geom_point() +
  theme_light() +
  theme(plot.title = element_text(face='bold'),
        legend.position = 'top') +
  labs(title='PCA by Tumour grade', 
       x = paste0('PC1 (', exp.var[1], '%)'), 
       y=paste0('PC2 (', exp.var[2], '%)'),
       color='Tumor Grade')
```



```{r}
ggplot(pca$x, aes(PC1, PC2, color=factor(meta_data$tumor_class))) +
  geom_point() +
  theme_light() +
  theme(plot.title = element_text(face='bold'),
        legend.position = 'right') +
  labs(title='PCA by Tumour Type', 
       x = paste0('PC1 (', exp.var[1], '%)'), 
       y=paste0('PC2 (', exp.var[2], '%)'),
       color='Tumor Type')
```



```{r}
var_expression <- apply(expr_data, 1, var)
```



```{r}
cbind(mean_expressions, var_expression) |>
  as.data.frame() |>
  pivot_longer(everything(), names_to = 'type', values_to = 'value') |>
  ggplot(aes(value)) +
  geom_density(color='steelblue') +
  facet_wrap(~type, scales='free') +
  theme_bw() +
  theme(strip.background = element_rect(fill='steelblue'),
        strip.text = element_text(color='white', face='bold'))
```

```{r}
summary(var_expression)
```

```{r}
sum(var_expression < quantile(var_expression, 0.25))
```

